\LARGE \bf Planning with Attitude

### 2020-09-09 paper/iros_abstract.tex
Begin adding IEEE version

%%%%%%%%%%%%%%%%%%%% author.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a proceedings volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{svproc}
\usepackage{url}
\def\UrlFont{\rmfamily}

@@

% Formatting
\usepackage{xcolor}
\usepackage{bm}  % for bold symbols 
\usepackage{booktabs}  % better tables
\usepackage{pifont}  % for x mark
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{wrapfig}
\usepackage{graphicx}

@@

%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ BEGIN DOCUMENT ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~%
\begin{document}
\mainmatter              % start of a contribution

\title{Planning and Control with Attitude}
\author{Brian Jackson \and Zachary Manchester}

\authorrunning{Jackson and Manchester}
\institute{Stanford University, 496 Lomita Mall, Stanford, CA, USA,\\
\email{\{bjack205,zacm\}@stanford.edu}}

\maketitle

\begin{abstract}
Planning and controlling trajectories for floating-base robotic systems that experience large attitude changes is challenging due to the nontrivial group structure of 3D rotations. 
This paper introduces an accessible and unified approach for tracking control and optimization-based planning on the space of rotations based on vector calculus and linear algebra. The methodology is used to derive an extension of the Linear-Quadratic Regulator (LQR) to systems with arbitrary attitudes, which we call Multiplicative LQR (MLQR). 
We compare MLQR to a state-of-the-art geometric tracking controller designed specifically for quadrotors and demonstrate superior tracking and robustness. 
An iterative variant of MLQR is compared to existing methods on useveral constrained trajectory optimization problems, again demonstrating improved runtime performance. %to optimize trajectories for a variety of robotic systems, we provide benchmark comparisons between several of the most common attitude representations used for motion planning and control and find that the combination of unit quaternion state variables with Rodrigues parameters to represent attitude errors provides an excellent combination of performance and algorithmic simplicity. 
\keywords{motion planning and control, quaternions, optimal control, linear quadratic regulator}
\end{abstract}

\section{Introduction}
    Many useful robotic systems---including quadrotors, airplanes, satellites, autonomous underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional translations and rotations as part of their normal operation. 
    While representing translations is straightforward and intuitive, effectively representing the nontrivial group structure of 3D rotations has been a topic of study for many decades. Although we can intuitively deduce that rotations are three-dimensional, a globally non-singular three-parameter representation of the space of rotations does not exist \cite{stuelpnagel1964parametrization}. As a result, when parameterizing rotations, we must either a) pick a three-parameter representation and deal with discontinuities, or b) pick a higher-dimensional representation and deal with constraints between the parameters. While simply representing attitude is nontrivial, generating and tracking motion plans for floating-base systems is an even more challenging problem.

    Early work on control problems involving the rotation group dates back to the 1970s, with extensions of linear control theory to spheres \cite{Brockett1973} and $SO(3)$ \cite{Baillieul1978}. Effective attitude tracking controllers have been developed for satellites \cite{wie1985quaternion}, quadrotors \cite{Fresk2013,Liu2015,lee2010geometric,Johnson2005,watterson2020control,mellinger2011minimum}, and a 3D inverted pendulum \cite{Chaturvedi2009}  using various methods for calculating three-parameter attitude errors.

    More recently, these ideas have been extended to trajectory generation \cite{Zefran1998}, sample-based motion planning \cite{Zefran1999,Kuffner2004}, and optimal control. Approaches to optimal control on attitude problems include analytical methods applied to satellites \cite{Spindler1998}, discrete mechanics \cite{Kobilarov2011,Kobilarov2014,Lee2008}, a combination of sampling-based planning and constrained trajectory optimization for satellite formations \cite{Garcia2005, Aoude2008}, projection operators \cite{Saccon2013}, or more general theory for optimization on manifolds \cites{watterson2018trajectory}. Nearly all of these methods rely heavily on principles from differential geometry and Lie group theory; however, despite these works, many recent papers in the robotics community continue to apply traditional methods for motion planning and control with no regard for the group structure of rigid body motion \cite{Alothman2016,deCrousaz2015,Williams2017,Geisert2016}.

    In this paper, we make a departure from previous approaches to geometric planning and control that rely heavily on ideas and notation from differential geometry, and instead use only basic mathematical tools from linear algebra and calculus that should be familiar to most roboticists. Similar to \cite{Mandic2011,Xu2016}, in Sec. \ref{sec:Quaternion_Calculus} we introduce a quaternion differential calculus, but take a significantly simpler and more general approach that unifies both planning and control, enabling straight-forward adaptation of existing algorithms to systems with quaternion states. To make this concrete, in Sec. \ref{sec:MLQR} we apply our approach to derive an extension to the linear-quadratic regulator (LQR) that we call multiplicative LQR (MLQR), which is the control dual to the multiplicative extended Kalman filter (MEKF) from the state estimation literature \cite{markley2014fundamentals}. In Sec. \ref{sec:experiments} we provide several simulation results demonstrating the application of our method to tracking control and constrained trajectory optimization.
    In summary, our contributions include:
    %\vspace{-10pt}
    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on standard vector calculus and linear algebraic operations
        \item Derivation of multiplicative LQR, an adaptation of LQR to the control of systems with quaternion states
        \item A set of benchmark motion-planning and trajectory-tracking problems in which we compare our approach to existing methods commonly used in the robotics community to account for 3D rotations
    \end{itemize}

\section{Background}
    We begin by defining some useful conventions and notation. Attitude is defined as the rotation from the robot's body frame to a global inertial frame. We also define gradients to be row vectors, that is, for $f(x) : \R^n \to \R$, $\pdv{f}{x} \in \R^{1\times n}$.

@@

        We will sometimes find it helpful to create a quaternion with zero scalar part from a vector $r \in \R^3$. We denote this operation as,
        \begin{equation}
            \hat{r} = H r \equiv \begin{bmatrix} 0 \\ I_3 \end{bmatrix} r.
        \end{equation}
        Unit quaternions rotate a vector through the operation $\hat{r}' = \q \otimes \hat{r} \otimes \q^*$. This can be equivalently expressed using matrix multiplication as
        \begin{align} 
            r' &= H^T L(q) R(q)^T H r = A(q)r , \label{eq:quaternion_rotation} %\\ 
               %&= H^T L(q) L(Hr) T q = H^T R(Hr) R(q)^T q %\label{eq:quaternion_rotation2}
        \end{align}
        where $A(q)$ is the rotation matrix in terms of the elements of the quaternion \cite{markley2014fundamentals}.

    % \todo{Merge 3.1 and 3.2. Start with unit quaternions (maybe put some of the section 4 material here), then define RPs and MRPs with the simple equations relating them to the quaternion. Keep it high level and cite references for details.}

    % \subsection{Rodrigues Parameters}
    %     Rodrigues parameters, also known as Gibbs vectors or the Cayley transformation of quaternion, are a computationally-efficient 3-parameter represention of rotations, obtained by projecting onto the plane tangent to the quaternion identity.
    %     Their definition (up to the scaling factor) is given in Table \ref{tab:retraction_maps} as the Cayley Map. 
    %     Unlike quaternions and MRPs, they do not have a sign ambiguity.

    %  \subsection{Modified Rodrigues Parameters}
    %     Modified Rodrigues Parameters (MRPs) are the newest parameter representation and can be considered a stereographic projection of the quaternion. 
    %     They are defined (up to the scaling factor) in Table \ref{tab:retraction_maps}. 
    %     MRPs have gained a lot of attention in the aerospace community since their singularity happens at \ang{360} and are a minimal 3-parameter representation. 
    %     %\cite{markley2014fundamentals} is an excellent resource for gaining further understanding of both Rodrigues parameters and MRPs. While predominantly only seen in the aerospace community, MRPs have recently been introduced to other fields such as computer graphics \cite{Terzakis2018}.

    \subsection{Rigid Body Dynamics} \label{sec:rigidbody_dynamics}
        In the current work we will restrict our focus to rigid bodies moving freely in 3D space. That is, we consider systems with dynamics of the following form:
        \begin{equation} \label{eq:rigid_body_dynamics}
            x = \begin{bmatrix} r \\ R \\ v \\ \omega \end{bmatrix}, \quad 
            \dot{x} = \begin{bmatrix} 
                v \\ 
                \texttt{kinematics}(R,\omega) \\ 
                \frac{1}{m} F_G(x,u) \\ 
                J^{-1}(\tau_L(x,u) - \omega \times J \omega) 
            \end{bmatrix}
        \end{equation}
        where $x$ and $u$ are the state and control vectors, $r \in \R^3$ is the position, $R \in SO(3)$ is the attitude, $v \in \R^3$ is the linear velocity, and $\omega \in \R^3$ is the angular velocity. $m \in \R$ is the mass, $J \in \R^{3\times3}$ is the inertia matrix, $F_G(x,u) \in \R^3$ are the forces in the global frame, $\tau_L(x,u)$ are the moments in the local (body) frame, and $\texttt{kinematics}$ are the kinematics for the chosen attitude representation. The kinematics for unit quaternions are
        \begin{equation}
            \dot{q} = \half \q \otimes \hat{\omega} = \half L(q) H \omega.
        \end{equation}
    %     \todo{Get rid of this table. Give the kinematics for each representation in line with its description above. Skip what you can and cite.}
    %     \begin{table}
    % 		\centering
    % 		\caption{Attitude Representations}
    % 		\begin{tabular}{lll}
    % 			\toprule
    % 			\textbf{Name} & \textbf{Kinematics} & \textbf{Singularity} \\
    % 			\midrule
    % 			Rotation Matrix & $R \skewmat{\omega_L}$ & None \\
    % 			Quaternions & $\half q \hat{\omega_L}$ & None \\
    % 			Euler Angles & see \cite{michael2010grasp} & $\theta \text{ (pitch)} = \ang{90}$ \\
    % 		    Rodrigues Param & $\half(I_3 + \skewmat{g} + g g^T) \omega$ & $\ang{180}$ \\
    % 		    MRP & $\frac{1+\norm{p}^2}{4}\left(I_3 + 2\frac{(\skewmat{p})^2 + \skewmat{p}}{1+\norm{p}^2}\right)\omega_L$ & \ang{360} \\
    % 			\toprule
    % 		\end{tabular}
    % 		\label{tab:attitude_kinematics}
    % 	\end{table}

    % \subsection{Trajectory Optimization}
    % \todo{I would try to get rid of this section}
    %     Trajectory optimization is a powerful framework for controlling complicated robotic systems. The value of trajectory optimization lies primarily in its generality, allowing it to be applied to a very broad class of dynamical systems. One of the objectives of this paper is to present a clear and performant method for solving trajectory optimization problems for rigid body motions, i.e. $SE(3)$. Trajectory optimization methods solve some variant of the following problem:
    %     \begin{mini}[3]
    %     	{x_{0:N},u_{0:N-1}}{\ell_N(x_N) + \sum_{k=0}^{N-1} \ell_k(x_k,u_k,\Delta t) }{}{}
    %     	\addConstraint{x_{k+1}}{=f(x_k,u_k,\Delta t), \; k = 1,...,N\text-1}
    %     	\addConstraint{g_k(x_k,u_k)}{\leq 0,\; \forall k}
    %     	\addConstraint{h_k(x_k,u_k)}{=0,\; \forall k}
    %     	\label{opt:discrete_trajopt},
    %     \end{mini}
    %     where $x_k \in \R^n, u_k \in \R^m$ are the state and control values at time step $k$, $N$ is the total number of knot points, $\ell_k$ is the stage cost function, $\ell_N$ is the terminal cost, and $f(x,u,\Delta t)$ are the discretized dynamics with step length $\Delta t$.

    %     There are many methods for solving trajectory optimization problems. Iterative LQR (iLQR) is a common method that has gain a lot of attention in recent years, mainly for it's speed and relative simplicity \cite{li2004iterative}. In the current work, we extend our implementation as part of the ALTRO solver detailed in \cite{howell2019altro}.

@@

    \begin{figure}
        \centering
        % \includegraphics[height=6cm]{figures/tangent_plane.png}
        \includegraphics[height=5cm]{figures/tangent_plane.tikz}
        \caption{When linearizing about a point $q$ on an sphere $\mathbb{S}^{n-1}$ in n-dimensional space, the tangent space $T$ is a plane living in $\R^{n-1}$, illustrated here with $n=3$. Therefore, when linearizing about a unit quaternion $q \in \Q$, the space of differential rotations lives in $\R^3$.}
        \label{fig:tangent_plane}
    \end{figure}

        Derivatives consider the effect an infinitesimal perturbation to the input has on an infinitesimal perturbation to the output. For vector spaces, the composition of the perturbation with the nominal value is simple addition and the infinitesimal perturbation lives in the same space as the original vector. For unit quaternions, however, neither of these are true; instead, they compose according to \eqref{eq:quat_mult}, and infinitesimal unit quaternions are (to first order) confined to a 3-dimensional plane tangent to $\Q$ (see Fig. \ref{fig:tangent_plane}).
        %, and are related through a non-linear mapping $\varphi : \R^3 \to \Q$ )
        The fact that differential unit quaternions are three-dimensional should make intuitive sense: Rotations are inherently three-dimensional and differential rotations should live in the same space as angular velocity, i.e. $\R^3$.

        There are many possible three-parameter representations for small rotations in the literature. Many authors use the exponential map \cite{Baillieul1978,Zefran1998,Lee2008,Saccon2013,Sola2017,Fan2016,watterson2018trajectory}, while others have used the Cayley map (also known as Rodrigues parameters) \cite{Kobilarov2011,Kobilarov2014}, Modified Rodrigues Parameters (MRPs) \cite{Terzakis2018}, or the vector part of the quaternion \cite{Fresk2013}. 
        We choose Rodrigues parameters \cite{markley2014fundamentals} because they are computationally efficient and do not inherit the sign ambiguity associated with unit quaternions. The mapping between a vector of Rodrigues parameters $\phi \in \R^3$ and a unit quaternion $q$ is known as the Cayley map:
        \begin{equation} \label{eq:cayley}
            q = \varphi(\phi) = \frac{1}{\sqrt{1 + \norm{\phi}^2}} \begin{bmatrix} 1 \\ \phi \end{bmatrix}.
        \end{equation}
        We will also make use of the inverse Cayley map:
        \begin{equation}
            \phi = \varphi^{-1}(q) = \frac{q_v}{q_s}.
        \end{equation}

    \subsection{Jacobian of Vector-Valued Functions}
        When taking derivatives with respect to quaternions, we must take into account both the composition rule and the nonlinear
        mapping between the space of unit quaternions and our chosen three-parameter error representation. 
        % Before demonstrating how this is done, let's consider the real-valued case we should already be familiar with. For some function $y = r(x) : \R^n \to \R^p$ with a differential value $\delta x \in \R^n$ applied to the input:
        % \begin{align}
        %     y + \delta y = r(x+\delta x) &\approx r(x) + \nabla r(x) \delta x,
        %     % &= \lim_{\delta x \to 0} x + \pdv{r}{x} \pdv{r + \delta x}{\delta x} = r + \pdv{r}{x} 
        % \end{align}
        % we calculate $\nabla r$ by taking the derivative with respect $\delta x$, as $\delta x$ approaches zero:
        % \begin{equation}
        %     \nabla r(x) = \lim_{\delta x \to 0} \pdv{r}{x} \pdv{(x + \delta x)}{\delta x} = \lim_{\delta x \to 0} \pdv{r}{x} I = \pdv{r}{x}
        % \end{equation}
        Let $\phi \in \R^3$ be a differential rotation applied to a function with quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(L(q) \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} L(q) H := \pdv{h}{q} G(q) = \pdv{h}{q} \begin{bmatrix} -q_v^T \\ sI_3 + \skewmat{q_v} \end{bmatrix}
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is the \textit{attitude Jacobian}, which essentially becomes a ``conversion factor'' allowing us to apply results from standard vector calculus to the space of unit quaternions. This form is particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the same (up to a constant scaling factor) for any choice of three-parameter attitude representation.

    \subsection{Hessian of Scalar-Valued Functions}
	    If the output of $h$ is a scalar ($p = 1$), then we can find its Hessian by taking the Jacobian of \eqref{eq:quat_gradient} with respect to $\phi$ using the product rule, again evaluated at $\phi = 0$:
	   % \begin{equation} \label{eq:hess1}
	   %     \nabla^2 h(q) = \lim_{\phi\to0} \left(L(q) \pdv{\varphi}{q} \right)^T \pdv[2]{h}{q} \left(L(q) \pdv{\varphi}{\phi} \right)+ \pdv{\phi}  \left(G(L(q) \varphi(\phi))^T \pdv{h}{q}^T \right) .
	   % \end{equation}
	   % Expanding the second term in \eqref{eq:hess1} and simplifying results in:
	    \begin{equation} \label{eq:quat_hessian}
            \nabla^2 h(q) = G(q)^T \pdv[2]{h}{q} G(q) + I_3 \pdv{h}{q}q,
	    \end{equation}
	    where the second term comes from the second derivative of $\varphi(\phi)$. Similar to $G(q)$, this ends up being the same (up to a scaling factor) for any choice of three-parameter attitude representation.

@@

    % \subsection{The Attitude Jacobian}
    %     We now derive the unit quaternion \textit{attitude Jacobian} that maps infinitesimal rotations into changes in the attitude quaternion. Consider the fairly trivial function $g(q,\phi) = L(q) \varphi(\phi)$, which simply composes a small rotation $\phi$ with a nominal attitude $q$, analagous to $x + \delta x$ in a vector space. Taking the derivative with respect to $\phi$ and evaluating it at $\phi=0$, we find the following Jacobian matrix, which we denote $G(q)$:
    %     \begin{equation} \label{eq:differential_jacobian}
    %         \pdv{f}{\phi}\Big|_{\phi=0} = L(q) \pdv{\varphi}{\phi}\Big|_{\phi=0} = L(q) H = \begin{bmatrix} -q_v^T \\ q_s I + \skewmat{q_v} \end{bmatrix} \equiv G(q)
    %     \end{equation}
    %     \todo{I still don't like this...} By differentiating through the inverse of $f$, $f^{-1}(q,q') = \varphi^{-1}(L(q')^T q)$, and evaluating at the unit quaternion identity, we
    %     can similarly find that
    %     \begin{equation}
    %         \pdv{f^{-1}}{q}\Big|_{q=I_q} = H^T L^T(q') = G(q')^T
    %     \end{equation}
    %     The attitude Jacobian essentially becomes a ``conversion factor'' allowing us to apply results from standard vector calculus to the space of unit quaternions. 
    %     As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the same (up to a constant scaling factor) for any choice of three-parameter attitude representation.

>>>>

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out if you need a4paper
\IEEEoverridecommandlockouts                              % This command is only needed if 
                                                          % you want to use the \thanks command
\overrideIEEEmargins                                      % Needed to meet printer requirements.

@@

% Formatting
\usepackage{xcolor}
\usepackage{bm}  % for bold symbols 
\usepackage{booktabs}  % better tables
\usepackage{pifont}  % for x mark
\usepackage{graphicx}

@@

\title{\LARGE \bf
Planning with Attitude
}

\author{Brian Jackson$^1$ and Zachary Manchester$^1$%
    \thanks{
        $^1$Robotics Institute, 
        Carnegie Mellon University, 
        5000 Forbes Ave, Pittsburgh, PA, USA
    }
}

\begin{document}
\maketitle

\section{INTRODUCTION}
    Many useful robotic systems---including quadrotors, airplanes, satellites, autonomous 
    underwater vehicles, and quadrupeds---can perform arbitrarily large three-dimensional 
    translations and rotations as part of their normal operation. 
    While representing translations is straightforward and intuitive, effectively 
    representing the nontrivial group structure of 3D rotations has been a topic of study 
    for many decades. Although we can intuitively deduce that rotations are three-dimensional, 
    a globally non-singular three-parameter representation of the space of rotations does 
    not exist \cite{stuelpnagel1964parametrization}. As a result, when parameterizing rotations, 
    we must either a) pick a three-parameter representation and deal with discontinuities, 
    or b) pick a higher-dimensional representation and deal with constraints between the 
    parameters. While simply representing attitude is nontrivial, generating and tracking 
    motion plans for floating-base systems is an even more challenging problem.

    Early work on control problems involving the rotation group dates back to the 1970s, 
    with extensions of linear control theory to spheres \cite{Brockett1973} and $SO(3)$ 
    \cite{Baillieul1978}. Effective attitude tracking controllers have been developed for 
    satellites \cite{wie1985quaternion}, quadrotors \cite{Fresk2013,Liu2015,lee2010geometric,
    Johnson2005,watterson2020control,mellinger2011minimum}, and a 3D inverted pendulum 
    \cite{Chaturvedi2009}  using various methods for calculating three-parameter attitude 
    errors.

    More recently, these ideas have been extended to trajectory generation \cite{Zefran1998}, 
    sample-based motion planning \cite{Zefran1999,Kuffner2004}, and optimal control. 
    Approaches to optimal control on attitude problems include analytical methods applied to
    satellites \cite{Spindler1998}, discrete mechanics \cite{Kobilarov2011,Kobilarov2014,
    Lee2008}, a combination of sampling-based planning and constrained trajectory 
    optimization for satellite formations \cite{Garcia2005, Aoude2008}, projection operators 
    \cite{Saccon2013}, or more general theory for optimization on manifolds 
    \cites{watterson2018trajectory}. 
    Nearly all of these methods rely heavily on principles from differential geometry and 
    Lie group theory; however, despite these works, many recent papers in the robotics 
    community continue to apply traditional methods for motion planning and control with no 
    regard for the group structure of rigid body motion \cite{Alothman2016,deCrousaz2015,
    Williams2017,Geisert2016}.

    In this paper, we make a departure from previous approaches to geometric planning and 
    control that rely heavily on ideas and notation from differential geometry, 
    and instead use only basic mathematical tools from linear algebra and calculus that 
    should be familiar to most roboticists. 
    Similar to \cite{Mandic2011,Xu2016}, in Sec. \ref{sec:Quaternion_Calculus} we introduce 
    a quaternion differential calculus, but take a significantly simpler and more general 
    approach that unifies both planning and control, enabling straight-forward adaptation of 
    existing algorithms to systems with quaternion states. 
    To make this concrete, in Sec. \ref{sec:MLQR} we apply our approach to derive an 
    extension to the linear-quadratic regulator (LQR) that we call multiplicative LQR (MLQR), 
    which is the control dual to the multiplicative extended Kalman filter (MEKF) from the 
    state estimation literature \cite{markley2014fundamentals}. 
    In Sec. \ref{sec:experiments} we provide several simulation results demonstrating the 
    application of our method to tracking control and constrained trajectory optimization.
    In summary, our contributions include:

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on standard vector calculus and linear algebraic operations
        \item Derivation of multiplicative LQR, an adaptation of LQR to the control of systems with quaternion states
        \item A set of benchmark motion-planning and trajectory-tracking problems in which we compare our approach to existing methods commonly used in the robotics community to account for 3D rotations
    \end{itemize}

\section{Background}

    We begin by defining some useful conventions and notation. 
    Attitude is defined as the rotation from the robot's body frame to a global inertial 
        frame. 
    We also define gradients to be row vectors, that is, for 
        $f(x) : \R^n \to \R$, $\pdv{f}{x} \in \R^{1\times n}$.

@@

        We will sometimes find it helpful to create a quaternion with zero scalar part from 
        a vector $r \in \R^3$. We denote this operation as,
        \begin{equation}
            \hat{r} = H r \equiv \begin{bmatrix} 0 \\ I_3 \end{bmatrix} r.
        \end{equation}
        Unit quaternions rotate a vector through the operation 
        $\hat{r}' = \q \otimes \hat{r} \otimes \q^*$. 
        This can be equivalently expressed using matrix multiplication as
        \begin{align} 
            r' &= H^T L(q) R(q)^T H r = A(q)r , \label{eq:quaternion_rotation}
        \end{align}
        where $A(q)$ is the rotation matrix in terms of the elements of the quaternion 
        \cite{markley2014fundamentals}.

    \subsection{Rigid Body Dynamics} \label{sec:rigidbody_dynamics}
        In the current work we will restrict our focus to rigid bodies moving freely in 3D 
        space. That is, we consider systems with dynamics of the following form:
        \begin{equation} \label{eq:rigid_body_dynamics}
            x = \begin{bmatrix} r \\ R \\ v \\ \omega \end{bmatrix}, \quad 
            \dot{x} = \begin{bmatrix} 
                v \\ 
                \texttt{kinematics}(R,\omega) \\ 
                \frac{1}{m} F_G(x,u) \\ 
                J^{-1}(\tau_L(x,u) - \omega \times J \omega) 
            \end{bmatrix}
        \end{equation}
        where $x$ and $u$ are the state and control vectors, $r \in \R^3$ is the position, 
        $R \in SO(3)$ is the attitude, $v \in \R^3$ is the linear velocity, and 
        $\omega \in \R^3$ is the angular velocity. $m \in \R$ is the mass, 
        $J \in \R^{3\times3}$ is the inertia matrix, $F_G(x,u) \in \R^3$ are the forces in the 
        global frame, $\tau_L(x,u)$ are the moments in the local (body) frame, and 
        $\texttt{kinematics}$ are the kinematics for the chosen attitude representation. The 
        kinematics for unit quaternions are
        \begin{equation}
            \dot{q} = \half \q \otimes \hat{\omega} = \half L(q) H \omega.
        \end{equation}

@@

    \begin{figure}
        \centering
        \includegraphics[height=5cm]{figures/tangent_plane.tikz}
        \caption{
            When linearizing about a point $q$ on an sphere $\mathbb{S}^{n-1}$ in 
            n-dimensional space, the tangent space $T$ is a plane living in $\R^{n-1}$, 
            illustrated here with $n=3$. Therefore, when linearizing about a unit 
            quaternion $q \in \Q$, the space of differential rotations lives in $\R^3$.
        }
        \label{fig:tangent_plane}
    \end{figure}

        Derivatives consider the effect an infinitesimal perturbation to the input has on
        an infinitesimal perturbation to the output. For vector spaces, the composition
        of the perturbation with the nominal value is simple addition and the
        infinitesimal perturbation lives in the same space as the original vector. For
        unit quaternions, however, neither of these are true; instead, they compose
        according to \eqref{eq:quat_mult}, and infinitesimal unit quaternions are (to
        first order) confined to a 3-dimensional plane tangent to $\Q$ (see Fig.
        \ref{fig:tangent_plane}).

        %, and are related through a non-linear mapping $\varphi : \R^3 \to \Q$ )
        The fact that differential unit quaternions are three-dimensional should make
        intuitive sense: Rotations are inherently three-dimensional and differential
        rotations should live in the same space as angular velocity, i.e. $\R^3$.

        skinny dog runs after skinny cat

        There are many possible three-parameter representations for small rotations in
        the literature. Many authors use the exponential map \cite{Baillieul1978,
        Zefran1998, Lee2008, Saccon2013, Sola2017, Fan2016, watterson2018trajectory},
        while others have used the Cayley map (also known as Rodrigues parameters)
        \cite{Kobilarov2011, Kobilarov2014}, Modified Rodrigues Parameters (MRPs)
        \cite{Terzakis2018}, or the vector part of the quaternion \cite{Fresk2013}.
        We choose Rodrigues parameters \cite{markley2014fundamentals} because they are
        computationally efficient and do not inherit the sign ambiguity associated with
        unit quaternions. The mapping between a vector of Rodrigues parameters $\phi \in
        \R^3$ and a unit quaternion $q$ is known as the Cayley map: \begin{equation}
        \label{eq:cayley}
            q = \varphi(\phi) = \frac{1}{\sqrt{1 + \norm{\phi}^2}} \begin{bmatrix} 1 \\ \phi \end{bmatrix}.
        \end{equation}
        We will also make use of the inverse Cayley map:
        \begin{equation}
            \phi = \varphi^{-1}(q) = \frac{q_v}{q_s}.
        \end{equation}

    \subsection{Jacobian of Vector-Valued Functions}
        When taking derivatives with respect to quaternions, we must take into account
        both the composition rule and the nonlinear mapping between the space of unit
        quaternions and our chosen three-parameter error representation.

        Let $\phi \in \R^3$ be a differential rotation applied to a function with
        quaternion inputs $y = h(q): \Q \to \R^p$, such that
        \begin{equation} \label{eq:vector_function}
            y + \delta y = h(L(q) \varphi(\phi)) \approx h(q) +  \nabla h(q) \phi.
        \end{equation}
        We can calculate the Jacobian $\nabla h(q) \in \R^{p \times 3}$ by
        differentiating \eqref{eq:vector_function} with respect to $\phi$, evaluated at
        $\phi = 0$:
        \begin{equation} \label{eq:quat_gradient}
            \nabla h(q) = \pdv{h}{q} L(q) H := \pdv{h}{q} G(q) 
                        = \pdv{h}{q} \begin{bmatrix} 
                            -q_v^T \\ 
                            sI_3 + \skewmat{q_v}
                        \end{bmatrix}
        \end{equation}
        where $G(q) \in \R^{4 \times 3}$ is the \textit{attitude Jacobian}, which
        essentially becomes a ``conversion factor'' allowing us to apply results from
        standard vector calculus to the space of unit quaternions. This form is
        particularly useful in practice since $\pdv*{h}{q} \in \R^{p \times 4}$ can be
        obtained using finite difference or automatic differentiation.
        As an aside, although we have used Rodrigues parameters, $G(q)$ is actually the
        same (up to a constant scaling factor) for any choice of three-parameter attitude
        representation.

    \subsection{Hessian of Scalar-Valued Functions}
	    If the output of $h$ is a scalar ($p = 1$), then we can find its Hessian by
	    taking the Jacobian of \eqref{eq:quat_gradient} with respect to $\phi$ using the
        product rule, again evaluated at $\phi = 0$:

	    \begin{equation} \label{eq:quat_hessian}
            \nabla^2 h(q) = G(q)^T \pdv[2]{h}{q} G(q) + I_3 \pdv{h}{q}q,
	    \end{equation}
	    where the second term comes from the second derivative of $\varphi(\phi)$.
	    Similar to $G(q)$, this ends up being the same (up to a scaling factor) for any
        choice of three-parameter attitude representation.

@@

\section{Multiplicative LQR} \label{sec:MLQR}
    Leveraging the methods from the previous section, we derive multiplicative LQR
    (MLQR), a variant of LQR that correctly accounts for the group structure of
    rotations. For concreteness, we consider a system with rigid body dynamics, as
    presented in Sec. \ref{sec:rigidbody_dynamics}, and design a controller to stabilize
    the system about a dynamically feasible discretized reference trajectory
    $\bar{x}_{0:N}, \bar{u}_{0:N}$ with $N$ time steps.
    We begin by linearizing the dynamics about the reference trajectory using \eqref{eq:quat_jacobian}. Our linearized error dynamics become
    \begin{equation} \label{eq:linearized_dynamics}
        \delta x_{k+1} = A_k \delta x_k + B_k \delta u_k 
    \end{equation}
    where \begin{equation}
        \begin{aligned}
            A_k = E(\bar{x}_{k+1})^T \pdv{f}{x}|_{\bar{x}_k,\bar{u}_k} E(\bar{x}_k), \\
            B_k = E(\bar{x}_{k+1})^T \pdv{f}{u}|_{\bar{x}_k,\bar{u}_k},
        \end{aligned}
    \end{equation}
    and $\delta x_k \in \R^{12}$ and $E(x_k) \in \R^{12 \times 13}$ are the state error and state error Jacobian:
    \begin{equation} \label{eq:state_error}
        \setlength\arraycolsep{1pt}
        \delta x_k = \begin{bmatrix} 
            r_k - \bar{r}_k \\ \varphi^{-1}(\bar{\q}_k^{-1} \otimes \q_k) \\ v_k - \bar{v}_k \\ \omega_k - \bar{\omega}_k 
        \end{bmatrix}\!, \;
        E(x) = \begin{bmatrix}
            I_3 & & & \\
            & G(q) & & \\
            & & I_3 & \\
            & & & I_3 \\
        \end{bmatrix}\!.
    \end{equation}

    With our linearized system, we can apply the traditional LQR Riccati recursion,
    \begin{multline}
         P_k = W_k +  A_k^T P_{k+1} A_k \\
         - A_k^T P_{k+1}^T B_k (R_k + B_k^T P_{k-1} B_k)^{-1} B_k^T P_{k+1} A_k ,
    \end{multline}
    which minimizes the quadratic cost function
    \begin{equation}
        \half \sum_{k=0}^N \delta x_k^T Q_k \delta x_k + \delta u_k^T R_k \delta u_k
    \end{equation}
    of the state \emph{error}, where $Q_k \in \R^{12 \times 12}$ is the weight matrix on
    the state \emph{error}, and $R_k \in \R^m$ is the weight matrix on the controls. From
    this, we get our non-linear feedback policy
    \begin{equation} \label{eq:mlqr_control}
        u = K_k \delta x_k + \bar{u}_k.
    \end{equation}
    where $\delta x_k$ is computed online using the non-linear error function
    \eqref{eq:state_error} and the linear feedback gain $K_k$ is calculated using the
    standard LQR formula:
    \begin{equation} \label{eq:LQR_gain}
         K_k := -(R_k + B_k^T P_{k+1} B_k)^{-1} B_k^T P_{k+1} A_k.
    \end{equation}
